---
title: "Component Biomass"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Component Biomass}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ForestBiomass)
```

`treeCalc()` returns total above-ground biomass by summing four components:
wood (stem), bark, branches, and foliage. Each component can also be computed
independently. This is useful when you need allocation patterns, want to
exclude a component for a particular tree type (e.g., no foliage on a snag),
or need per-component carbon accounting.

## Individual component functions

All four functions share the same argument structure as `treeCalc()`.

```{r}
wood <- woodCalc(
  data    = sample_trees, eval = "ung_2",
  species = "SPECIES", dbh = "DBH", height = "HEIGHT",
  output  = "biomass", decay = FALSE
)

bark <- suppressMessages(barkCalc(
  data    = sample_trees, eval = "ung_2",
  species = "SPECIES", dbh = "DBH", height = "HEIGHT",
  output  = "biomass", decay = FALSE
))

branch <- suppressMessages(branchCalc(
  data       = sample_trees, eval = "ung_2",
  species    = "SPECIES",    dbh = "DBH", height = "HEIGHT",
  crown_cond = "CROWN_COND", output = "biomass", decay = FALSE
))

foliage <- suppressMessages(foliageCalc(
  data       = sample_trees, eval = "ung_2",
  species    = "SPECIES",    dbh = "DBH", height = "HEIGHT",
  crown_cond = "CROWN_COND", output = "biomass", decay = FALSE
))
```

Summing the four components reproduces the `treeCalc()` result exactly:

```{r}
agb <- suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", crown_cond = "CROWN_COND",
           output = "biomass", decay = FALSE)
)
all.equal(wood + bark + branch + foliage, agb)
```

Component proportions vary by species and tree size:

```{r}
total <- wood + bark + branch + foliage
data.frame(
  species = sample_trees$SPECIES,
  wood_pct    = round(wood    / total * 100, 1),
  bark_pct    = round(bark    / total * 100, 1),
  branch_pct  = round(branch  / total * 100, 1),
  foliage_pct = round(foliage / total * 100, 1)
)
```

## Adjusting for partial bark loss

`barkCalc()` accepts an optional `rem_bark` column (percentage remaining,
0-100). This scales bark biomass proportionally and is useful for trees with
beetle kill, fire scorch, or logging damage.

```{r}
sample_trees$REM_BARK <- c(100, 100, 80, 60, 20, 100)

full_bark <- suppressMessages(
  barkCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", output = "biomass", decay = FALSE)
)
partial_bark <- suppressMessages(
  barkCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", rem_bark = "REM_BARK",
           output = "biomass", decay = FALSE)
)

data.frame(
  species  = sample_trees$SPECIES,
  rem_bark = sample_trees$REM_BARK,
  full_kg  = round(full_bark,    2),
  adj_kg   = round(partial_bark, 2)
)
```

## Jenkins component ratios

`jenkinsRatio()` decomposes total above-ground biomass into four component
proportions using Jenkins et al. (2003) equations. It requires only DBH and
species — no height or crown condition. Species are classified as softwood or
hardwood internally.

```{r}
ratios <- jenkinsRatio(sample_trees, dbh = "DBH", species = "SPECIES")
str(ratios)
```

Each element is a vector of ratios (one per tree). The Jenkins equations are
fit independently per component, so the four components sum to approximately 1
rather than exactly 1:

```{r}
wood_type <- SPECIES_CLASS$WOOD_TYPE[
  match(sample_trees$SPECIES, SPECIES_CLASS$SPECIES)
]

data.frame(
  species     = sample_trees$SPECIES,
  wood_type   = wood_type,
  foliage     = round(ratios$foliage,      3),
  coarse_root = round(ratios$coarse_roots, 3),
  stem_bark   = round(ratios$stem_bark,    3),
  stem_wood   = round(ratios$stem_wood,    3),
  check_sum   = round(ratios$foliage + ratios$coarse_roots +
                        ratios$stem_bark + ratios$stem_wood, 3)
)
```

Multiply the ratios by a Jenkins total biomass estimate to get component values.
Note that `coarse_root` here is a Jenkins-derived estimate and uses different
equations than `rootCalc()`, which applies Li et al. (2003):

```{r}
jenkins_total <- suppressMessages(
  treeCalc(sample_trees, eval = "jenkins", species = "SPECIES",
           dbh = "DBH", output = "biomass", decay = FALSE)
)

data.frame(
  species     = sample_trees$SPECIES,
  stem_wood   = round(ratios$stem_wood    * jenkins_total, 1),
  stem_bark   = round(ratios$stem_bark    * jenkins_total, 1),
  foliage     = round(ratios$foliage      * jenkins_total, 1),
  coarse_root = round(ratios$coarse_roots * jenkins_total, 1)
)
```
