---
title: "Getting Started with ForestBiomass"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ForestBiomass}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(ForestBiomass)
```

ForestBiomass calculates above-ground and below-ground tree biomass using
published allometric equations. This vignette walks through the most common
workflow: estimating total above-ground biomass with `treeCalc()`, choosing
output units, applying decay corrections, and chaining into `rootCalc()` for
root biomass.

## Your data

`treeCalc()` requires a data frame with one row per tree. The package includes
`sample_trees`, a six-tree dataset covering common BC species:

```{r}
sample_trees
```

The required columns are:

| Column | Content | Notes |
|--------|---------|-------|
| Species | NFI code (e.g., `"PSEU_MEN"`) | See `?UNG_1` for valid Ung codes |
| DBH | Diameter at breast height (cm) | Required for all equation sets |
| Height | Total tree height (m) | Required for `ung_2` and `lambert_2` |
| Appearance | Appearance code (1-7) | Required when `decay = TRUE` |
| Crown condition | Crown condition code (1-5) | Required for Ung and Lambert pathways |

## Estimating total above-ground biomass

`treeCalc()` with `eval = "ung_2"` (the default) uses Ung et al. (2008) equations
for BC species, fitting biomass as a function of both DBH and height.

```{r}
agb <- treeCalc(
  data       = sample_trees,
  eval       = "ung_2",
  species    = "SPECIES",
  dbh        = "DBH",
  height     = "HEIGHT",
  appearance = "APPEARANCE",
  crown_cond = "CROWN_COND",
  output     = "biomass",
  decay      = TRUE
)
agb
```

Each value is the total above-ground biomass for one tree in kilograms.
`decay = TRUE` applies a species-specific reduction for standing dead or downed
wood (Harmon et al., 2011). For live-tree inventories, use `decay = FALSE`.

## Output options

The `output` argument controls units and whether the result is carbon:

```{r}
# Per-tree biomass (kg) — default
suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "biomass", decay = FALSE)
)

# Per-tree carbon (kg) — biomass × 0.5
suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "carbon", decay = FALSE)
)

# Stand-level biomass (Mg/ha) — requires plot_radius
suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "biomass Mg/ha",
           decay = FALSE, plot_radius = 11.28)
)

# Stand-level carbon (Mg/ha)
suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "carbon Mg/ha",
           decay = FALSE, plot_radius = 11.28)
)
```

The `plot_radius` argument controls the per-hectare expansion factor:

| Plot radius (m) | Trees per hectare |
|-----------------|-------------------|
| 3.99 | 200 |
| 5.64 | 100 |
| 7.98 | 50 |
| 11.28 | 25 |

## Choosing an equation set

Five `eval` options are available:

| `eval` | Source | Inputs needed | Best for |
|--------|--------|---------------|----------|
| `"ung_1"` | Ung et al. 2008 | DBH | BC — DBH only |
| `"ung_2"` | Ung et al. 2008 | DBH + height | BC — DBH and height (default) |
| `"lambert_1"` | Lambert et al. 2005 | DBH | Eastern Canada — DBH only |
| `"lambert_2"` | Lambert et al. 2005 | DBH + height | Eastern Canada — DBH and height |
| `"jenkins"` | Jenkins et al. 2003 | DBH | Any region — DBH only |

`"ung_1"` and `"ung_2"` use the same 17 BC species. Use `"lambert_1"` or
`"lambert_2"` for Eastern Canada species (see `?LAMBERT_1`). The `"jenkins"`
pathway covers all species in `JENKINS_GROUPS` and requires no height or crown
condition measurement.

## Decay class reduction

When `decay = TRUE` (the default), a species-specific decay class reduction
factor is applied from Harmon et al. (2011). Trees with appearance code 1
(freshly dead / recently fallen) receive no reduction; higher codes apply
increasing reductions. Live trees should use `decay = FALSE`.

```{r}
no_decay   <- suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "biomass", decay = FALSE)
)
with_decay <- suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "biomass", decay = TRUE)
)

data.frame(
  species    = sample_trees$SPECIES,
  appearance = sample_trees$APPEARANCE,
  no_decay   = round(no_decay,   1),
  with_decay = round(with_decay, 1),
  pct_reduction = round((1 - with_decay / no_decay) * 100, 1)
)
```

## Using the Jenkins pathway

When height or crown condition are unavailable, `eval = "jenkins"` estimates
total above-ground biomass directly from DBH alone:

```{r}
treeCalc(
  data       = sample_trees,
  eval       = "jenkins",
  species    = "SPECIES",
  dbh        = "DBH",
  appearance = "APPEARANCE",
  output     = "biomass",
  decay      = TRUE
)
```

## Estimating root biomass

`rootCalc()` applies Li et al. (2003) root equations to a pre-computed
above-ground biomass column. It requires the above-ground biomass to be in
Mg/ha (or kg with a `plot_radius` for conversion).

```{r}
sample_trees$AGB <- suppressMessages(
  treeCalc(sample_trees, eval = "ung_2", species = "SPECIES",
           dbh = "DBH", height = "HEIGHT", appearance = "APPEARANCE",
           crown_cond = "CROWN_COND", output = "biomass Mg/ha",
           decay = TRUE, plot_radius = 11.28)
)

rootCalc(
  data      = sample_trees,
  biomass   = "AGB",
  species   = "SPECIES",
  units     = "Mg/ha",
  root_type = "total",
  output    = "biomass Mg/ha"
)
```

Fine and coarse root fractions can be returned separately:

```{r}
fine   <- suppressMessages(
  rootCalc(sample_trees, biomass = "AGB", species = "SPECIES",
           units = "Mg/ha", root_type = "fine")
)
coarse <- suppressMessages(
  rootCalc(sample_trees, biomass = "AGB", species = "SPECIES",
           units = "Mg/ha", root_type = "coarse")
)
total  <- suppressMessages(
  rootCalc(sample_trees, biomass = "AGB", species = "SPECIES",
           units = "Mg/ha", root_type = "total")
)

data.frame(
  species = sample_trees$SPECIES,
  fine    = round(fine,   3),
  coarse  = round(coarse, 3),
  total   = round(total,  3)
)
```
